---
title: "walkshed"
author: "Victor"
date: "2024-07-17"
output: html_document
---
#Install Packages

```{r setup, include=FALSE}
libs <- c('viridis', 'sf','stars','tidyverse','raster',
          'terra','ggplot2','osmdata','httr','leaflet','gt','mapview','webshot',
          'htmltools','htmlwidgets','rnaturalearth','rnaturalearthdata') #needed libraries

installed_libs <- libs %in% rownames(installed.packages())

if(any(installed_libs == F)) {
  install.packages(libs[!installed_libs])
} 

invisible(lapply(libs,library,character.only = T))

```

#load data
```{r}
nigeria <- ne_countries(scale = 'medium',country = 'nigeria', returnclass = 'sf') #nigeria sf



nga_pop <- rast("\\Users\\User\\Documents\\GIS\\walkshed\\NGA_ppp_v2c_2020_UNadj.tif")#population data
```

#Extract Train stations within Nigeria

```{r}
get_osm_stations <- function() {
  bbox <- sf::st_bbox(nigeria) 
  train_stations <- bbox |>
    opq() |>
    add_osm_feature(
      key = "railway",
      value = 'station'
    ) |>
    osmdata::osmdata_sf()
  
  return(train_stations)
} #create a function to extract train stations in Nigeria



train_stations_main <- get_osm_stations() #Assign to a new object

train_stations_main <- train_stations_main$osm_points #assign points as new column

```

#Extract Railway lines within Nigeria
```{r}
railway_tags <- c('light_rail','monorail','rail') #OSM tags for railway types
get_osm_railways <- function() {
  bbox <- st_bbox(nigeria)
  rails <- bbox |>
    opq() |>
    add_osm_feature(
      key = "railway",
      value = railway_tags
    ) |>
    osmdata::osmdata_sf()
  
  return(rails)
}

nigeria_rails <- get_osm_railways()


nigeria_rails <-nigeria_rails$osm_lines

```

#extract lines of interest 
```{r}
standard_track_gauge_light_rail<-nigeria_rails %>% dplyr::filter(gauge %in% c(1435,1067)) #filter railway lines

```

#extract train stations along lines of interest
```{r}
standard_track_gauge_utm <- st_transform(standard_track_gauge_light_rail,crs=32633) #change coordinate reference system to utmWGS84
train_stations_utm <- st_transform(train_stations_main,crs=32633)

rail_buffer <- st_buffer(standard_track_gauge_utm,dist=1000)

rail_buff_union <- st_union(rail_buffer)

stations_online <- sf::st_intersection(train_stations_utm,rail_buff_union) #extract stations that fall within buffer

stations_oi <- c('Abuja Metro','Kukwaba II','Gbazango')

stations_interest <-stations_online %>% dplyr::filter(name %in% stations_oi | wikidata == 'Q110556255') 
stations_interest <- st_transform(stations_interest,crs=4326) #transform to a geographic coordinate system

stations_interest_coord <- as.data.frame(st_coordinates(stations_interest))
stations_interest_coord$name <- stations_interest$name
stations_interest_coord <- stations_interest_coord[c('name','X','Y')] #reorder columns

stations <- data.frame(name =c("Wupa","Gwagwa","Deidei","Kagini","Bassanjiwa","Kukwaba I","Stadium","Airport"),
                       X = c( 7.3948998,7.2851857,7.2872407,7.2921787,7.2824573,7.4410513,7.4517158,7.2723953),
                       Y=c(9.0246825,9.0898296, 9.1061639, 9.1246465, 9.0136807, 9.0402465,9.0459078,9.0067099)) 


stations <- bind_rows(stations_interest_coord,stations)

stations <- st_as_sf(stations,coords = c("X","Y"), crs = 4326)

```

#create a buffer of 800m and 400m for walkshed
```{r}
station_buffer<- st_buffer(stations,dist=800)



station_buffer_400 <- st_buffer(stations, dist = 400)

```

#Populate a column with respective distances of each buffer for the purpose of displaying on legend
```{r}
station_buffer <- station_buffer %>%  mutate(distance_t = 800)


station_buffer_400 <- station_buffer_400 %>%  mutate(distance_t = 400)

```

#Bind rows for walkshed
```{r}
buffer_merged <- bind_rows(station_buffer,station_buffer_400)


buffer_merged$distance_t <- factor(buffer_merged$distance_t)

buffer_merged <-buffer_merged %>% 
  dplyr::select(name,distance_t)

buffer_grouped <-  buffer_merged %>% 
  dplyr::group_by(distance_t) 

```

#carry out zonal statistics for number of people within each walk shed
```{r}
Zonal_sta <-terra::zonal(nga_pop,vect(station_buffer),sum,na.rm = T)

station_buffer$pop_dens <- Zonal_sta$NGA_ppp_v2c_2020_UNadj

```

#Arrange names of station in order of route
```{r}
custom_order <- c("Abuja Metro", "Stadium", "Kukwaba I", "Kukwaba II", "Wupa", 
                  "Idu", "Gwagwa", "Deidei", "Kagini", "Gbazango", 
                  "Bassanjiwa", "Airport")

station_buffer$name <- factor(station_buffer$name, levels = custom_order)

station_buffer <- station_buffer %>% arrange(name) %>% 
  dplyr::select(name,pop_dens)

station_buffer$pop_dens <- round(station_buffer$pop_dens)

```

#Make a table showing the population of people within each walk shed
```{r}
station_buffer_df <- st_drop_geometry(station_buffer)


catchment_population <- gt(station_buffer_df) %>% 
  cols_label(name =md("*Station Name*"),
             pop_dens = "Catchment Population") %>% 
  gt::tab_header(title = "Abuja Rail Mass Transit",
                 subtitle = md("*How many people can take a 5 minute walk to a station?*") )%>% 
  tab_options(data_row.padding = px(6),
              heading.align = 'left',
              column_labels.background.color = '#CBF3F0',
              row_group.background.color = '#CBF3F0',
              heading.title.font.size = px(26),
  )  %>% 
  tab_source_note(md("*Author:@VictorA47*"))

```

#convert table to html to be used in leaflet
```{r}
catchment_population_html <- as.character(catchment_population %>% as_raw_html())

pal_fact <- leaflet::colorFactor("viridis",
                                 domain = buffer_grouped$distance_t,
                                 reverse = T)

```

#Create a walk shed for all stations on the light rail line

```{r}
walkshed <- leaflet(buffer_grouped) %>% 
  
  addPolygons(fill = T,
              stroke = T,
              weight = -3,
              fillColor = ~pal_fact(buffer_grouped$distance_t) ,
              fillOpacity = .5,
              label = ~buffer_grouped$name)%>% 
  
  addProviderTiles('OpenStreetMap.Mapnik') %>% 
  addLegend("bottomright",
            values = buffer_grouped$distance_t,
            labels = buffer_grouped$distance_t,
            title = "Area Within Walking Distance(m)",
            pal = pal_fact) %>% 
  addControl(
    html = catchment_population_html,
    position = "topright",
    className = "info legend"
  )

walkshed
```


